An Experimental Study of Sorting and Branch Prediction

### 2007-01-29 docs/branch_paper/paper.tex
Formatting changes for paper.tex and paper.bib, updated README, removed unused directory.

\author{
PAUL BIGGAR\footnote[1]{PUT IN IRCSET GRANT DETAILS}, NICHOLAS NASH\footnotemark[1] KEVIN WILLIAMS\footnotemark[1] and DAVID GREGG
\\University of Dublin, Trinity College
}

@@

Another type of instruction whose cost can vary dramatically is the
conditional branch. Modern pipelined processors depend on \emph{branch
prediction} for much of their performance.  If the direction of a
conditional branch is correctly predicted ahead of time, the cost of
the conditional branch may be as little as the cost of, say, an
integer add instruction. If, on the other hand, the branch is
mispredicted the processor must flush its pipeline, and restart
from the correct target of the branch. This cost is typically a large
multiple of the cost of executing a correctly-predicted branch. For
example, Intel Pentium 4 processors \cite{Intel248966-010,Intel249438-01} have pipelines of up 
to 31 stages,
meaning that a branch misprediction will cost around 30 cycles.  In
contrast, on the same processor the cost of a memory access that
misses the first-level cache and hits the second level cache is 23
cycles. Fortunately, the branches in most programs are very
predictable, so branch mispredictions are usually rare. Indeed, prediction
accuracies of greater than 90\% are typical \cite{Uht+97} with the
best predictors.

@@

Processors do not generally provide access to the behaviour of their caches or branch predictors. However, a software processor simulator can be used to 
provide measurements 
of cache and branch prediction behaviour.
Our branch prediction and caching results were generated using the SimpleScalar PISA simulator version 3 \cite{SimpleScalarTutorialv4}. 
We used \texttt{sim-cache} and 
\texttt{sim-bpred} to generate results for caching and branch prediction characteristics respectively.
We used a variety of cache configurations; generally speaking we used an 8 KB level 1 data cache, 8 KB level 1 instruction cache,
and a shared 2 MB instruction and data level 2 cache, all with 32 byte cache lines. 
We also experimented with both direct mapped and fully associative caches. When presenting results
where the cache configuration is important, we will describe the precise configuration used. We present
results here only for direct mapped caches. The results for fully associative caches are similar \cite{BiggarGregg05}.

@@

\begin{verbatim}
   item = a[i];
   while(item < a[i - 1]) 
   {
     a[i] = a[i - 1];
     i--;
   }
   a[i] = item;
\end{verbatim}
On the $i^{th}$ iteration \texttt{a[0..i - 1]} is sorted, allowing
\texttt{a[i]} to be placed in the correct position as shown in
the inner-loop above. On average, insertion sort performs approximately
$n^2/4$ comparisons and $n^2/4$ assignments in its inner-loop. Thus insertion sort
performs around half as many branches as selection sort.

@@

\begin{verbatim}
for(j = 0; j < n - i - 1; j++)
{
    if(a[j + 1] < a[j])
       swap(a[j + 1], a[j]);
}
\end{verbatim} 
In this code $i$ is the outer-loop iteration counter, and begins at 0 and counts
towards $n - 1$. Bubble sort uses
information yielded by the comparisons which selection
sort simply discards. After a full iteration of the inner-loop shown larger
elements have moved to the right, closer to their final location.

@@

To understand the high misprediction rate of bubble sort we define $Q^k_j$
as the probability that $j^{th}$ inner-loop comparison branch is taken on the $k^{th}$ 
outer-loop iteration, $1 \leq k < n$. It turns out that

\begin{equation}
\label{bubble_equation}
Q^k_j = 
\left\{ 
\begin{array}{ll}
  \frac{j}{j + k} & \quad \mbox{if $j \leq n - k$}\\
  0 & \quad \mbox{otherwise}\\ 
\end{array} 
\right. 
\end{equation}

\noindent
This is a result of the fact that the $j^{th}$ comparison branch of the inner-loop
is taken only if \texttt{a[j]} is not larger than all of \texttt{a[0..j - 1]}.
For this
to be the case, the $(j + 1)^{th}$ branch in the previous outer-loop iteration must have been taken,
since otherwise the previous iteration of the outer-loop left the maximum of \texttt{a[0..j - 1]} in \texttt{a[j]}.
The probability that the $(j + 1)^{th}$ branch of the
previous outer-loop iteration is taken is $Q^{k - 1}_{j + 1}$. Given that the $(j + 1)^{th}$ branch of
the previous outer-loop iteration was taken, the probability that \texttt{a[j]} is
not larger than all of \texttt{a[0..j - 1]} is $1 - 1/(j + 1)$, thus,

\[
Q^k_j = Q^{k-1}_{j+1}\left(1 - \frac{1}{j + 1} \right)
\]

\noindent
The bases $Q^{1}_j = 1 - 1 / (j + 1)$ for $1 \leq j < n$ and $Q^1_n = 0$ give the solution shown in Equation \ref{bubble_equation}.
It is easy to see that in the first outer-loop iteration ($k = 1$) the comparison branch is very likely to be taken, whereas in the last
outer-loop iteration ($k = n - 1$), the comparison branch is very unlikely to be taken. Moving between these two extremes causes the intermediate
branches to be unpredictable. The remainder of this section examines this phenomenon in more detail.

@@

\begin{eqnarray*}
Q^k_{avg} &=& \frac{1}{n - k} \sum_{j = 1}^{n - k} \frac{j}{j + k} \\
          &=& 1 + \frac{k}{n - k}(H_k - H_n)\\
%  &=& \frac{n - k - k(H_n - H_k)}{n - k}\\
         &\approx& 1 + \frac{k}{n - k}\ln\frac{k}{n}
\end{eqnarray*}

@@

Shellsort \cite{Shell59} was the first in-place sorting algorithm
with time complexity better than $O(n^2)$.
Algorithms like selection and bubble sort use each comparison
to resolve at most one inversion (an inversion is 
a pair of keys in the input which appear in the wrong order) in the input, and
it is well known that this restricts their average case running
time to be in $O(n^2)$. Shellsort uses each comparison to
resolve more than one inversion, giving better performance.

@@

\begin{figure}
\centering
\begin{tabular}{cc}
\includegraphics[width=0.5\textwidth]{plots/shellsort_branch_misses.eps} & \includegraphics[width=0.5\textwidth]{plots/shell_heap_cache_misses.eps} \\
(a) & (b) \\
\end{tabular}
\caption{(a) Shows the average branches misses per key for shellsort using a bimodal and two-level adaptive predictor, both having 4096 table entries.
On this occasion, the two-level adaptive predictor out-performs the simpler bimodal predictor. \texttt{sim-bpred}
was used to simulate these branch predictors.
(b) Shows that the average number of cache misses per key of our basic heapsort implementation
is higher than for our basic shellsort implementation. Shellsort incurs only a small number of cache misses
because on average a key moves just 0.9744 places per iteration. These results use \texttt{sim-cache} to simulate
an 8KB direct mapped cache with 32 byte cache lines.}
\label{shellsort_results_figs}
\end{figure}

@@

One might expect shellsort's cache performance to be poor, since it repeatedly accesses
distantly separated keys in its input while the increments are large. Each such access
could potentially cause a cache miss. However, the fact that on average 
keys move just 0.9744 places per iteration
implies that large numbers of references in distant parts
of the input (which could lead to cache misses) are not generally necessary.
In fact, the number of cache misses for our shellsort implementations, although significant,
is smaller than for our basic heapsort implementation as Figure \ref{shellsort_results_figs}(b)
shows. For our input sizes, which were up to $2^{22}$ keys, our shellsort implementations were on
average about 50\% faster than the best of our heapsort implementations (see Section \ref{heapsort} for
further details on our heapsort implementations).

@@

@@

When using a $k$-ary heap, finding the maximum (or minimum) child at a particular node causes the $i^{th}$ comparison branch to be taken
with probability $1/(1 + i)$. The comparisons behave just like those in selection sort (see Section \ref{selection_sort}), and are therefore quite
predictable.  For our 8-heap we unrolled the
loop to find the minimum child into 7 separate branches. 
Figure \ref{heapsort_results}(d) shows the behaviour of these 7 branches, for an unrolled loop.
As expected, the later branches are gradually taken less often on average, with a corresponding increase in their predictability.
Unrolling this loop
not only removes the loop over-head, but is also experimentally observed to improve the average prediction accuracy by about 2\%. 
The average steady state predicitability in the unrolled case, is given by

@@

We also analysed the branch prediction behaviour in several
cache conscious mergesort variations, using the work of LaMarca and Ladner \citeyear{LaMarca97} as a reference.
Mergesort is not in-place\footnote{Worst case linear time in-place merging is possible but is generally inefficient in practice, see for example Katajainen 
\textit{et al} \citeyear{Katajainen+96}.}
and uses an auxiliary array 
for
its merge operation. With a cache of size $C$, the largest input we can sort in the cache therefore has size $C/2$. To avoid conflict misses, the
arrays should be allocated so that the cache line index of the start of the source array is $C/2$ from the cache line index of the start of the destination array. 
This can
be done for both the level 1 and level 2 cache.

@@

\begin{figure}
\centering
\begin{tabular}{cc}
\includegraphics[width=0.5\textwidth]{plots/mergesort_branches_per_key.eps} & \includegraphics[width=0.5\textwidth]{plots/mergesort_branch_misses.eps} \\
(a) & (b) \\
\includegraphics[width=0.5\textwidth]{plots/mergesort_level_2_misses.eps} & \includegraphics[width=0.5\textwidth]{plots/mergesort_cycle_counts.eps} \\
(c) & (d) \\
\end{tabular}
\caption{(a) Shows the branches per key executed by our mergesort variations. The sudden increase in the number of executed branches for multi-mergesort
results from the use of the multi-way merge when the data no longer fits within the cache. (b) Shows the number of mispredictions per key, all predictors
had 4096 table entries. Bimodal predictors were used for base and tiled mergesort. These results were generated using \texttt{sim-bpred}, the
two-level adaptive predictor had a 10-bit history register.
(c) Shows the level 2 cache misses per key, base mergesort has by far the worst cache behaviour, while multi-mergesort
has slightly better cache performance than tiled mergesort. These results were generated by using \texttt{sim-cache} to simulate a
direct mapped 2MB cache with 32-byte cache lines. Finally (d) shows the cycles per key of our algorithms measured using Pentium 4 hardware performance
counters. Tiled mergesort performs the best out of the algorithms, despite its inferior cache performance compared to multi-mergesort.}
\label{mergesort_branch_results}
\end{figure}

@@

The implementations of multi-quicksort have the best cache performance of the quicksort variations, as Figure \ref{quicksort_plots}(a) shows.
However, the multi-quicksort implementations also introduce many extra instructions per key as Figure \ref{quicksort_plots}(b) shows. Naturally, the 
sequentially searched version introduces more extra instructions than the binary searched version. However, as described in the previous paragraph,
the branches introduced by binary search are highly unpredictable. As a result, multi-quicksort using binary search has more branch mispredictions than the
sequential searched version (see Figure \ref{quicksort_plots}(c)), consequently, despite having identical cache performance, binary searched multi-quicksort
performs worse than the sequentially searched multi-quicksort, as shown in Figure \ref{quicksort_plots}(d). Of course, on very large data sets (larger than the $2^{22}$ keys we experimented with), the 
reduced instruction count of binary searched multi-quicksort would eventually allow it to out-perform sequentially searched multi-quicksort.

@@

In work concurrent and independent of our own, Kaligosi and Sanders \citeyear{Kaligosi+06} 
have investigated the effect of the choice of pivot on the performance of quicksort. They found that
an artificially skewed pivot gives better performance than using a random pivot or the median-of-3,
due to the reduction in branch mispredictions. This performance increase is despite the increased instruction count
such an artificially skewed pivot causes, we describe this work in more detail in Section \ref{RelatedWork}.

@@

In our basic radix sort implementation, the first array is accessed 3 times and the second array 4 times, in each of the four
iterations over the four bytes in the 32-bit integers. This gives 20 accesses in total.
In memory-tuned radix sort each key of the two arrays is accessed only nine times in total.
At the start, it traverses the original array once to count the least significant bytes. It then iterates four times, positioning
the keys based on the previously established counts. These four iterations cause
a total of eight accesses, because each array is accessed once per iteration.

@@

Our radix sort implementation requires fewer processor cycles per key than even our base quicksort 
implementation as Figure \ref{radixsort_results_pictures}(a) shows. The number of level 2 cache misses is only slightly higher than we measured for
quicksort while the array fits inside the cache, presumably because radix sort is out-of-place.

Figure \ref{radixsort_results_pictures}(b) shows the level 2 cache misses for our basic and memory-tuned implementations.
We note that, when the array no longer fits inside the cache the memory-tuned implementation has approximately half the number of misses per key
(1.25) than the basic version (2.5). When the data no longer fits inside the cache there should be about one miss for each key per pass per cache line.
Each cache line can fit 8 keys, and thus we expect about 2 misses per key since there are 16 passes in total. The 2.5 misses per key which was measured
likely exceeds this because of the irregular fashion of the array indexing in each pass. Similary, in memory-tuned radix sort we expect about 1.125
misses per key, because there are 9 passes, again the measured 1.25 misses per key presumably exceeds this because
of the irregular array indexing.

@@

In  
preliminary experiments, we found that Sanders and Winkel's work is difficult 
to reproduce on a Pentium 4 based machine. The only predicated instruction
supported on the Pentium 4 is the conditional move instruction. We found that
rewriting
quicksort's loops so that they are suitable for predication
reduces their efficiency and also increases the number of simultaneously
live variables. Since the Pentium 4 has only eight general
purpose registers the number of simultaneously live variables exceeded
the number of available registers, reducing efficiency. In addition,
we also discovered that on the Pentium 4 conditional moves are
surprisingly expensive.

Mudge \textit{et al} \citeyear{Mudge+96} examine the behaviour of the $i$ and $j$
comparison branches in randomized quicksort, as an example to demonstrate a
general method for estimating the limit on the predictability of
branches in a program. They argue that the optimal
predictor for these branches would 
keep a running count of the total
number of array elements examined so far that are
greater than the pivot. If the majority of the elements examined so far
are greater than the pivot, the next element is predicted as greater
than the pivot, and vice versa. They estimate that this approach will give a
prediction accuracy of 75\%, which is close to our
result for these branches, using a bimodal predictor.
This shows that a bimodal predictor is close to optimal
for randomized quicksort's branches, since $P_{avg}$ of Section \ref{quicksort_branch_results_text}
is about 0.71.

>>>>

\author{
PAUL BIGGAR\footnote[1]{Supported by the Irish Research Council for Science,
Engineering and Technology (IRCSET).}, NICHOLAS NASH\footnotemark[1], KEVIN
WILLIAMS\footnote[2]{Supported by the Irish Research Council for Science,
Engineering and Technology (IRCSET) and IBM.}, and DAVID GREGG \\University of
Dublin, Trinity College
}

@@

Another type of instruction whose cost can vary dramatically is the conditional
branch. Modern pipelined processors depend on \emph{branch prediction} for much
of their performance.  If the direction of a conditional branch is correctly
predicted ahead of time, the cost of the conditional branch may be as little as
the cost of, say, an integer add instruction. If, on the other hand, the branch
is mispredicted the processor must flush its pipeline, and restart from the
correct target of the branch. This cost is typically a large multiple of the
cost of executing a correctly-predicted branch. For example, Intel Pentium 4
processors \cite{Intel248966-010,Intel249438-01} have pipelines of up to 31
stages, meaning that a branch misprediction will cost around 30 cycles.
Fortunately, the branches in most programs are very predictable, so branch
mispredictions are usually rare. Indeed, prediction accuracies of greater than
90\% are typical \cite{Uht+97} with the best predictors.

@@

In order to experiment with a variety of cache and branch prediction results we
used the SimpleScalar PISA processor simulator version 3
\cite{SimpleScalarTutorialv4}.  We used \texttt{sim-cache} and
\texttt{sim-bpred} to generate results for caching and branch prediction
characteristics respectively.

We used a variety of cache configurations; generally speaking we used an 8 KB
level 1 data cache, 8 KB level 1 instruction cache, and a shared 2 MB
instruction and data level 2 cache, all with 32 byte cache lines.  We also
experimented with both direct mapped and fully associative caches. When
presenting results where the cache configuration is important, we will describe
the precise configuration used. We present results here only for direct mapped
caches. The results for fully associative caches are similar, as we show in our
technical report \cite{BiggarGregg05}, which contains complete results for a
large number of branch predictor and cache configurations.

@@

\begin{verbatim}
   item = a[i];
   while(item < a[i - 1]) 
   {
     a[i] = a[i - 1];
     i--;
   }
   a[i] = item;
\end{verbatim}

On the $i^{th}$ iteration \texttt{a[0..i - 1]} is sorted, allowing \texttt{a[i]}
to be placed in the correct position as shown in the inner-loop above. On
average, insertion sort performs approximately $n^2/4$ comparisons and $n^2/4$
assignments in its inner-loop. Thus insertion sort performs around half as many
branches as selection sort.

@@

\begin{verbatim}
for(j = 0; j < n - i - 1; j++)
{
    if(a[j + 1] < a[j])
       swap(a[j + 1], a[j]);
}

\end{verbatim}

In this code $i$ is the outer-loop iteration counter, and begins at 0 and counts
towards $n - 1$. Bubble sort uses information yielded by the comparisons which
selection sort simply discards. After a full iteration of the inner-loop shown
larger elements have moved to the right, closer to their final location.

@@

To understand the high misprediction rate of bubble sort we define $Q^k_l$ as
the probability that $l^{th}$ inner-loop comparison branch is taken on the
$k^{th}$ outer-loop iteration, $1 \leq k < n$. It turns out that

\begin{equation}
\label{bubble_equation}
Q^k_l = 
\left\{ 
\begin{array}{ll}
  \frac{l}{l + k} & \quad \mbox{if $l \leq n - k$}\\
  0 & \quad \mbox{otherwise}\\ 
\end{array} 
\right. 
\end{equation}

\noindent
This is a result of the fact that the $l^{th}$ comparison branch of the
inner-loop is taken only if \texttt{a[l]} is not larger than all of
\texttt{a[0..l - 1]}.  For this to be the case, the $(l + 1)^{th}$ branch in the
previous outer-loop iteration must have been taken, since otherwise the previous
iteration of the outer-loop left the maximum of \texttt{a[0..l - 1]} in
\texttt{a[l]}.  The probability that the $(l + 1)^{th}$ branch of the
previous outer-loop iteration is taken is $Q^{k - 1}_{l + 1}$. Given that the
$(l + 1)^{th}$ branch of the previous outer-loop iteration was taken, the
probability that \texttt{a[j]} is not larger than all of \texttt{a[0..l - 1]} is
$1 - 1/(l + 1)$, thus,

\[
Q^k_l = Q^{k-1}_{l+1}\left(1 - \frac{1}{l + 1} \right)
\]

\noindent
The bases $Q^{1}_l = 1 - 1 / (l + 1)$ for $1 \leq l < n$ and $Q^1_n = 0$ give
the solution shown in Equation \ref{bubble_equation}.  It is easy to see that in
the first outer-loop iteration ($k = 1$) the comparison branch is very likely to
be taken, whereas in the last outer-loop iteration ($k = n - 1$), the comparison
branch is very unlikely to be taken. Moving between these two extremes causes
the intermediate branches to be unpredictable. The remainder of this section
examines this phenomenon in more detail.

@@

\begin{eqnarray*}
Q^k_{avg} &=& \frac{1}{n - k} \sum_{l = 1}^{n - k} \frac{l}{l + k} \\
          &=& 1 + \frac{k}{n - k}(H_k - H_n)\\
%  &=& \frac{n - k - k(H_n - H_k)}{n - k}\\
         &\approx& 1 + \frac{k}{n - k}\ln\frac{k}{n}
\end{eqnarray*}

@@

Shellsort \cite{Shell59} was the first in-place sorting algorithm with time
complexity better than $O(n^2)$.  Algorithms like selection and bubble sort use
each comparison to resolve at most one inversion (an inversion is a pair of keys
in the input which appear in the wrong order) in the input, and it is well known
that this restricts their average case running time to be in $O(n^2)$
\cite{KnuthVol3_98}. Shellsort uses each comparison to resolve more than one
inversion, giving better performance.

@@

@@

One might expect shellsort's cache performance to be poor, since it repeatedly
accesses distantly separated keys in its input while the increments are large.
Each such access could potentially cause a cache miss. However, the fact that on
average keys move just 0.9744 places per iteration implies that large numbers of
references in distant parts of the input (which could lead to cache misses) are
not generally necessary.  In fact, our shellsort implementation has better cache
performance in both the level 1 and level 2 cache than for our basic heapsort
implementation as Figure \ref{shellsort_results_figs}(b), and Figure
\ref{shellsort_results_figs}(c) shows.  Indeed, our shellsort implementation
performs significantly better than our basic heapsort implementation as Figure
\ref{shellsort_results_figs}(d) shows (see Section \ref{heapsort} for further
details on our heapsort implementations).

@@

\begin{figure}
\centering
\begin{tabular}{cc}
\includegraphics[width=0.5\textwidth]{plots/shellsort_branch_misses.eps} & \includegraphics[width=0.5\textwidth]{plots/shell_heap_cache_misses.eps} \\
(a) & (b) \\
\includegraphics[width=0.5\textwidth]{plots/shell_heap_L2_misses.eps} & \includegraphics[width=0.5\textwidth]{plots/shell_heap_cycles.eps} \\
(a) & (b) \\
\end{tabular}
\caption{(a) Shows the average branches misses per key for shellsort using a
bimodal and two-level adaptive predictor, both having 4096 table entries.  On
this occasion, the two-level adaptive predictor out-performs the simpler bimodal
predictor. \texttt{sim-bpred} was used to simulate these branch predictors.  (b)
and (c) show that our basic heapsort implementation has inferior cache
performance at both levels of the cache compared to our shellsort
implementation. Shellsort incurs only a small number of cache misses because on
average a key moves just 0.9744 places per iteration. These results use
\texttt{sim-cache} to simulate an 8KB direct mapped level 1 cache and a 2MB
direct mapped level 2 cache both with 32 byte cache lines. Finally (d) shows
that our shellsort implementation out-performs our heapsort implementation,
these results were measured using Pentium 4 hardware performance counters.}
\label{shellsort_results_figs}
\end{figure}

@@

When using a $k$-ary heap, finding the maximum (or minimum) child at a
particular node causes the $i^{th}$ comparison branch to be taken with
probability $1/(1 + i)$. The comparisons behave just like those in selection
sort (see Section \ref{selection_sort}), and are therefore quite predictable.
For our 8-heap we unrolled the loop to find the minimum child into 7 separate
branches.  Figure \ref{heapsort_results}(d) shows the behaviour of these 7
branches, for an unrolled loop.  As expected, the later branches are gradually
taken less often on average, with a corresponding increase in their
predictability.  Unrolling this loop not only removes the loop over-head, but is
also experimentally observed to improve the average prediction accuracy by about
2\% \cite{BiggarGregg05}.  The average steady state predicitability in the
unrolled case, is given by

@@

We also analysed the branch prediction behaviour in several cache conscious
mergesort variations using the work of LaMarca and Ladner \citeyear{LaMarca97}
as a reference, the details of our implementations differ slightly however
\cite{BiggarGregg05}.  Mergesort is not in-place\footnote{Worst case linear time
in-place merging is possible but is generally inefficient in practice, see for
example Katajainen \textit{et al} \citeyear{Katajainen+96}.} and uses an
auxiliary array for its merge operation. With a cache of size $C$, the largest
input we can sort in the cache therefore has size $C/2$. To avoid conflict
misses, the arrays should be allocated so that the cache line index of the start
of the source array is $C/2$ from the cache line index of the start of the
destination array.  This can be done for both the level 1 and level 2 cache.

@@

\begin{figure}
\centering
\begin{tabular}{cc}
\includegraphics[width=0.5\textwidth]{plots/mergesort_branches_per_key.eps} & \includegraphics[width=0.5\textwidth]{plots/mergesort_branch_misses.eps} \\
(a) & (b) \\
\includegraphics[width=0.5\textwidth]{plots/mergesort_level_2_misses.eps} & \includegraphics[width=0.5\textwidth]{plots/mergesort_cycle_counts.eps} \\
(c) & (d) \\
\end{tabular}
\caption{(a) Shows the branches per key executed by our mergesort variations.
The sudden increase in the number of executed branches for multi-mergesort
results from the use of the multi-way merge when the data no longer fits within
the cache. (b) Shows the number of mispredictions per key, all predictors had
4096 table entries. Bimodal predictors were used for base and tiled mergesort.
These results were generated using \texttt{sim-bpred}, the two-level adaptive
predictor had a 10-bit history register.  (c) Shows the level 2 cache misses per
key, base mergesort has by far the worst cache behaviour, while multi-mergesort
has slightly better cache performance than tiled mergesort. These results were
generated by using \texttt{sim-cache} to simulate a direct mapped 2MB cache with
32-byte cache lines. Finally (d) shows the cycles per key of our algorithms
measured using Pentium 4 hardware performance counters. Despite
multi-mergesort's superior cache performance, its heightened instruction count
and branch mispredictions mean that the less cache conscious tiled mergesort
performs the best out of the algorithms.}
\label{mergesort_branch_results}
\end{figure}

@@

The implementations of multi-quicksort have the best cache performance of the
quicksort variations, as Figure \ref{quicksort_plots}(a) shows.  However, the
multi-quicksort implementations also introduce many extra instructions per key
as is seen in Figure \ref{quicksort_plots}(b).  Naturally, the sequentially
searched version introduces more extra instructions than the binary searched
version. However, as described in the previous paragraph, the branches
introduced by binary search are highly unpredictable. As a result,
multi-quicksort using binary search has more branch mispredictions than the
sequential searched version (see Figure \ref{quicksort_plots}(c)), consequently,
despite having identical cache performance, binary searched multi-quicksort
performs worse than the sequentially searched multi-quicksort, as shown in
Figure \ref{quicksort_plots}(d). Of course, on very large data sets (larger than
the $2^{22}$ keys we experimented with), the reduced instruction count of binary
searched multi-quicksort would eventually allow it to out-perform sequentially
searched multi-quicksort.

@@

In work concurrent to and independent of our own, Kaligosi and Sanders
\citeyear{Kaligosi+06} have investigated the effect of the choice of pivot on
the performance of quicksort. They found that an artificially skewed pivot gives
better performance than using a random pivot or the median-of-3, due to the
reduction in branch mispredictions. This performance increase is despite the
increased instruction count such an artificially skewed pivot causes; we
describe this work in more detail in Section \ref{RelatedWork}.

@@

In our basic radix sort implementation, the first array is accessed 3 times and
the second array twice, in each of the four iterations over the four bytes in
the 32-bit integers. This gives 20 accesses in total.  In memory-tuned radix
sort each key of the two arrays is accessed only nine times in total.  At the
start, it traverses the original array once to count the least significant
bytes. It then iterates four times, positioning the keys based on the previously
established counts. These four iterations cause a total of eight accesses,
because each array is accessed once per iteration.

@@

Our radix sort implementation requires fewer processor cycles per key than even
our base quicksort implementation as Figure \ref{radixsort_results_pictures}(a)
shows.

Figure \ref{radixsort_results_pictures}(b) shows the level 2 cache misses for
our basic and memory-tuned implementations.  The number of level 2 cache misses
is only slightly higher than we measured for quicksort while the array fits
inside the cache, presumably because radix sort is out-of-place.  We note that
when the array no longer fits inside the cache the memory-tuned implementation
has approximately half the number of misses per key (1.25) than the basic
version (2.5). When the data no longer fits inside the cache there should be
about one miss for each key per pass per cache line.  Each cache line can fit 8
keys, and thus we expect the observed 2.5 misses per key since there are 20
passes in total.  likely exceeds this because of the irregular fashion of the
array indexing in each pass. Similary, in memory-tuned radix sort we expect
slightly below 1.25 misses per key since there are 9 passes in total.

@@

\comment{ In preliminary experiments, we found that Sanders and Winkel's work is
difficult to reproduce on a Pentium 4 based machine. The only predicated
instruction supported on the Pentium 4 is the conditional move instruction. We
found that rewriting quicksort's loops so that they are suitable for predication
reduces their efficiency and also increases the number of simultaneously live
variables.  Since the Pentium 4 has only eight general purpose registers the
number of simultaneously live variables exceeded the number of available
registers, reducing efficiency. In addition, we also discovered that on the
Pentium 4 conditional moves are surprisingly expensive.  } Mudge \textit{et al}
\citeyear{Mudge+96} examine the behaviour of the $i$ and $j$ comparison branches
in randomized quicksort, as an example to demonstrate a general method for
estimating the limit on the predictability of branches in a program. They argue
that the optimal predictor for these branches would keep a running count of the
total number of array elements examined so far that are greater than the pivot.
If the majority of the elements examined so far are greater than the pivot, the
next element is predicted as greater than the pivot, and vice versa. They
estimate that this approach will give a prediction accuracy of 75\%, which is
close to our result for these branches, using a bimodal predictor.  This shows
that a bimodal predictor is close to optimal for randomized quicksort's
branches, since $P_{avg}$ of Section \ref{quicksort_branch_results_text} is
about 0.71.
